Marginally-Stable Thermal Equilibria of Rayleigh-Bénard Convection

### 2021-05-27 publication_materials/ms_rbc.tex
amplitude equation arguments

To derive the quasilinear form, we posit that an arbitrary field $f$ can be represented as the sum of a mean profile (denoted by $\overline{f}$) and a perturbation function (denoted by $f'$).
\begin{align}
    \vec{u}(x, z, t) &= \vec{u'}(x, z, t) \label{EQ:reynolds_dc_u}\\
    &= u'(x, z, t)\hat{x} + w'(x, z, t)\hat{z} \\
    T(x, z, t) &= \overline{T}(z, t) + T'(x, z, t) \label{EQ:reynolds_dc_T}\\
    p(x, z, t) &= \overline{p}(z, t) +  p'(x, z, t) \label{EQ:reynolds_dc_p}.
\end{align}
where the mean-velocity components vanish due to incompressibility and symmetry. Perturbations are defined to have no horizontal-average
\begin{equation}
    \langle f'(x, z, t) \rangle_x \equiv \int_{0}^{L_x} f'(x, z, t) dx = 0.
\end{equation}
%Assuming the existance of nontrivial solutions, we also fix the volume-average over the entire domain
%\begin{equation}
%  \langle T'^2 \rangle_{\mathcal{D}} = 1
%\end{equation}
%thereby alleviating any ambiguity in $A(t)$. 
Substituting \eq{EQ:reynolds_dc_T} into \eq{EQ:motion3} and taking the horizontal-average reduces the system to a simple initial value problem (IVP) for $\overline{T}$
\begin{equation}
  \frac{\partial \overline{T}}{\partial t} + \frac{\partial}{\partial z} \langle w'T' \rangle_x = \mathcal{P} \frac{\partial^2 \overline{T}}{\partial z^2}, \label{EQ:T0_IVP}
\end{equation}
with associated boundary conditions $\overline{T}(-1/2, t) = 1/2$ and $\overline{T}(1/2, t) = -1/2$. It should be noted that we could obtain a similar IVP for $u$ by breaking reflection symmetry about the midplane $z = 0$ and considering some nontrivial mean horizontal flow $\overline{u}(z, t)$. However we must have $\overline{w}(z, t) = 0$ due to incompressibility.

@@

In his groundbreaking report \cite{Rayleigh_1916}, Lord Rayleigh observed that \eqss{EQ:linear1}{EQ:linear3} can be manipulated into a separable form with generalized solutions
\begin{align}
    w'(x, z, t) &= A\, \Re\left[W(z) \, e^{i(k_xx-st)}\right], \label{EQ:normal_modes1}\\ 
    u'(x, z, t) &= A\, \Re\left[U(z) \, e^{i(k_xx-st)}\right], \label{EQ:normal_modes2}\\ 
    T'(x, z, t) &= A\, \Re\left[\theta(z) \, e^{i(k_xx-st)}\right], \label{EQ:normal_modes3}\\ 
    p'(x, z, t) &= A\, \Re\left[P(z) \, e^{i(k_xx-st)}\right], \label{EQ:normal_modes4}
\end{align}
where $A$ is the (undetermined) mode amplitude, $s = \omega + i\sigma$ is the eigenvalue, and $k_x$ is constrained, by periodicity, to the countably infinite set (spectrum) of wavenumbers
\begin{align}
    k_x \in \left\{\frac{n\pi}{2} \, \big| \, n \in \mathbb{N}\right\}.
\end{align}
We normalize the eigenmodes to have
\begin{equation}
  \langle |\theta|^2 \rangle_{\mathcal{D}} = 1
\end{equation}
where $\langle \cdot \rangle_{\mathcal{D}}$ denotes the spatial mean over the entire domain.

@@

To initialize our calculations, we construct a marginally-stable initial temperature profile $\overline{T}(z, t=0)$ whose equation is given in Appendix~\ref{sec:initial_profile}. 
We use \eq{EQ:T0_IVP} to evolve $\overline{T}(z, t_0)$ into a new marginally-stable profile $\overline{T}(z, t_0 + \Delta t)$.
We use a second-order, two-stage IMEX Runge-Kutta method.
The eigenfunctions and amplitudes are assumed to be constant over the timestep.
We symmetrize $\overline{T}$ by setting the coefficients of its even Chebyshev basis functions to zero at the start of each timestep.
This ensures $\overline{T}(z) = -\overline{T}(-z)$.
On every timestep, we calculate the marginally stable eigenfunctions and their amplitude $A$.
It is essential that we pick the correct eigenfunction amplitude when calculating the advective term to maintain marginal stability.
We will now illustrate our method of finding the appropriate $A$ through an example.

@@

Consider a marginally-stable temperature profile $\overline{T}(z, t_0)$.
By definition, its maximum growth rate is 0.
Diffusing $\overline{T}(z, t_0)$ tends to increase its growth rates while ignoring the diffusive term and evolving according to advection tends to decrease its growth rates.
The amplitude $A$ must be selected such that these two influences are equal and opposite.
We can measure the effects of diffusion and advection on the maximum eigenvalue by solving two new IVPs
\begin{align}\label{eqn:diffusion_only}
    &\frac{\partial \overline{T}_{\rm{diff}}}{\partial t} = \mathcal{P} \frac{\partial^2 \overline{T}_{\rm{diff}}}{\partial z^2} \\
    &\frac{\partial \overline{T}_{\rm{adv}}}{\partial t} + \frac{\partial}{\partial z} \langle w'T' \rangle_x = 0 
\end{align}
where $\overline{T}_{\rm{diff}}$ and $\overline{T}_{\rm{adv}}$ denote the diffused and advected temperature profiles.
We initialize $\overline{T}_{\rm{diff}}$ and $\overline{T}_{\rm{adv}}$ with $\overline{T}(z, t_0)$, and evolve the equations to $t=t_0+\Delta t$.
Suppose $k_0$ is the wavenumber of the marginally-stable mode.
Then, using the diffused and advected profiles, we compute the eigenvalues $\sigma_{\rm{diff}}$ and $\sigma_{\rm{adv}}$ at $k_x = k_0$ and $t=t_0+\Delta t$.
A preliminary approximation $\Lambda^2_0$ of the appropriate amplitude $A^2$ is given by
\begin{equation}
    A^2 \approx \Lambda^2_0 \equiv -\frac{\sigma_{\rm{diff}}}{\sigma_{\rm{adv}}} \label{EQ:amp_approx}
\end{equation}
where the square in $A^2$ and $\Lambda^2_0$ is due to the quadratic nonlinearity of the eigenfunctions in the advective term and the minus sign is due to the fact that $\sigma_{\rm{adv}} < 0$ and $\sigma_{\rm{diff}} > 0$ in general.
We illustrate these trends in Figure~\ref{fig:iteration_spectra}. 
The subscript 0 denotes a preliminary estimate which we later refine.
In general, when $\Lambda^2_0 > 0$, the root-finding algorithm is stable.
The alternative can occur when the timestep approaches machine precision, in which case $\sigma_{\rm{diff}}$ and $\sigma_{\rm{adv}}$ are subject to floating point errors of comparable order to themselves.

Using the approximate amplitude $\Lambda_0^2$, the background temperature $\overline{T}(z, t_0)$ is then evolved to $t=t_0+\Delta t$ according to \eq{EQ:T0_IVP} and another eigenvalue solve is performed. 
This background temperature is typically close to, but not exactly, marginally stable.
We use Newton's method to find an amplitude $A^2$ such that the maximum growth rate is zero to within $10^{-9}$.
Marginally-stable modes do not oscillate in time, i.e. $\sigma = 0$ implies $\omega = 0$.
This agrees with the conventional notion of exchange of stabilities \cite{drazin_reid_2004}.
Crucially, we do not assume the $k_x$ of the marginally-stable mode is fixed.
In Section~\ref{sec:multiple_modes} we specify procedures for the treatment of multiple simultaneously marginal modes.

>>>>

To derive the quasilinear form, we posit that an arbitrary field $f$ can be represented as the sum of a mean profile (denoted by $\overline{f}$) and a perturbation function (denoted by $f'$).
\begin{align}
    \vec{u}(x, z, t) &= \vec{u'}(x, z, t) \label{EQ:reynolds_dc_u}\\
    &= u'(x, z, t)\hat{x} + w'(x, z, t)\hat{z} \\
    T(x, z, t) &= \overline{T}(z, t) + T'(x, z, t) \label{EQ:reynolds_dc_T}\\
    p(x, z, t) &= \overline{p}(z, t) +  p'(x, z, t) \label{EQ:reynolds_dc_p}.
\end{align}
where the mean-velocity components vanish due to incompressibility and symmetry. Perturbations are defined to have no horizontal-average
\begin{equation}
    \langle f'(x, z, t) \rangle_x \equiv \int_{0}^{L_x} f'(x, z, t) dx = 0.
\end{equation}
%Assuming the existance of nontrivial solutions, we also fix the volume-average over the entire domain
%\begin{equation}
%  \langle T'^2 \rangle_{\mathcal{D}} = 1
%\end{equation}
%thereby alleviating any ambiguity in $A(t)$. 
Substituting \eq{EQ:reynolds_dc_T} into \eq{EQ:motion3} and taking the horizontal-average reduces the system to a simple initial value problem for $\overline{T}$
\begin{equation}
  \frac{\partial \overline{T}}{\partial t} + \frac{\partial}{\partial z} \langle w'T' \rangle_x = \mathcal{P} \frac{\partial^2 \overline{T}}{\partial z^2}, \label{EQ:T0_IVP}
\end{equation}
with associated boundary conditions $\overline{T}(-1/2, t) = 1/2$ and $\overline{T}(1/2, t) = -1/2$. It should be noted that we could obtain a similar equation for $u$ by breaking reflection symmetry about the midplane $z = 0$ and considering some nontrivial mean horizontal flow $\overline{u}(z, t)$. However we must have $\overline{w}(z, t) = 0$ due to incompressibility.

@@

In his groundbreaking report \cite{Rayleigh_1916}, Lord Rayleigh observed that \eqss{EQ:linear1}{EQ:linear3} can be manipulated into a separable form with generalized solutions
\begin{align}
    w'(x, z, t) &= A\, \Re\left[W(z) \, e^{i(k_xx-st)}\right], \label{EQ:normal_modes1}\\ 
    u'(x, z, t) &= A\, \Re\left[U(z) \, e^{i(k_xx-st)}\right], \label{EQ:normal_modes2}\\ 
    T'(x, z, t) &= A\, \Re\left[\theta(z) \, e^{i(k_xx-st)}\right], \label{EQ:normal_modes3}\\ 
    p'(x, z, t) &= A\, \Re\left[P(z) \, e^{i(k_xx-st)}\right], \label{EQ:normal_modes4}
\end{align}
where $A$ is the (undetermined) mode amplitude, $s = \omega + i\sigma$ is the eigenvalue, and $k_x$ is constrained, by periodicity, to the countably infinite set (spectrum) of wavenumbers
\begin{align}
    k_x \in \left\{\frac{n\pi}{2} \, \big| \, n \in \mathbb{N}\right\}.
\end{align}
We normalize the eigenmodes to have
\begin{equation}
  \langle |\theta|^2 \rangle_{\mathcal{D}} = 1
\end{equation}
where $\langle \cdot \rangle_{\mathcal{D}}$ denotes the spatial mean over the entire domain.
Crucially, we emphasize that  \eqs{EQ:normal_modes1}{EQ:normal_modes3} are solutions to the linear problem.
In the quasilinear context, the background state and its eigenmodes vary in time.
By convention, our notation neglects this $t$-dependence in the eigenfunctions as they will always computed from the autonomous linear system.

@@

To initialize our calculations, we construct a marginally-stable initial temperature profile $\overline{T}(z, t=0)$ whose equation is given in Appendix~\ref{sec:initial_profile}. 
We use \eq{EQ:T0_IVP} to evolve $\overline{T}(z, t_0)$ into a new marginally-stable profile $\overline{T}(z, t_0 + \Delta t)$.
We use a second-order, two-stage IMEX Runge-Kutta method.
The eigenfunctions and amplitudes are assumed to be constant with respect to $t$ over the timestep.
We symmetrize $\overline{T}$ by setting the coefficients of its even Chebyshev basis functions to zero at the start of each timestep.
This ensures $\overline{T}(z) = -\overline{T}(-z)$.
On every timestep, we calculate the marginally stable eigenfunctions and their amplitude $A$.
It is essential that we pick the correct eigenfunction amplitude when calculating the advective term to maintain marginal stability.
We will now illustrate our method of finding the appropriate $A$ through an example.

@@

Consider a marginally-stable temperature profile $\overline{T}(z, t_0)$.
By definition, its maximum growth rate is 0.
Diffusing $\overline{T}(z, t_0)$ tends to increase its growth rates while ignoring the diffusive term and evolving according to advection tends to decrease its growth rates.
The amplitude $A$ must be selected such that these two influences are equal and opposite.
We can measure the effects of diffusion and advection on the maximum eigenvalue by solving two new initial value problems
\begin{align}\label{eqn:diffusion_only}
    &\frac{\partial \overline{T}_{\rm{diff}}}{\partial t} = \mathcal{P} \frac{\partial^2 \overline{T}_{\rm{diff}}}{\partial z^2} \\
    &\frac{\partial \overline{T}_{\rm{adv}}}{\partial t} + \frac{\partial}{\partial z} \Re\left[ W \theta^* \right] = 0 \label{eqn:advection_only}
\end{align}
where $\overline{T}_{\rm{diff}}$ and $\overline{T}_{\rm{adv}}$ denote the diffused and advected temperature profiles respectively. 
The advective term must be defined in this way because we do not yet know its amplitude $A^2$. 
In the original problem \eq{EQ:T0_IVP}, evolution due to advection is proportional to $A^2$.
That is not the case in \eq{eqn:advection_only}.

To approximate $A^2$, we assume the system is in a state of near-equilibrium, evolving on a slow time scale relative to diffusion or advection individually. 
If we initialize both \eqs{eqn:diffusion_only}{eqn:advection_only} with a marginally-stable\footnote[3]{Note that marginal stability differs from equilibrium in this quasilinear context because a marginally-stable state need not satisfy $\frac{\partial \overline{T}}{\partial t} = 0$.} temperature profile $\overline{T}$, the evolution of the background state due to diffusion must be approximately opposite the evolution due to advection on the fast time scale, i.e.
\begin{align}
    \overline{T}_{\rm{diff}} - \overline{T} &\approx -A^2(\overline{T}_{\rm{adv}} - \overline{T})
\end{align}
where the $A^2$ factor is meant to reconcile the different advective terms in \eqs{EQ:T0_IVP}{eqn:advection_only}.

This approximation is valid in equilibrium, but in general, we would obtain undesired $z$-dependence in $A^2$.
Though we might be tempted to take a norm, the best way to account for this $z$-dependence, is to make an analogous argument with respect to the growth rates
\begin{align}
    \sigma_{\rm{diff}} - \sigma &\approx -A^2(\sigma_{\rm{adv}} - \sigma).
\end{align}
where $\sigma_{\rm{diff}}$ and $\sigma_{\rm{adv}}$ are the growth rates computed from $\overline{T}_{\rm{diff}}$ and $\overline{T}_{\rm{adv}}$ respectively.
$\sigma$ vanishes due to marginal stability.
Later we will see that this approach can be easily generalized to account for multiple simultaneously marginal modes.

In practice, we initialize $\overline{T}_{\rm{diff}}$ and $\overline{T}_{\rm{adv}}$ with $\overline{T}(z, t_0)$, and evolve the equations to $t=t_0+\Delta t$.
Suppose $k_0$ is the wavenumber of the marginally-stable mode.
Then we compute $\sigma_{\rm{diff}}$ and $\sigma_{\rm{adv}}$ at $k_x = k_0$ and $t=t_0+\Delta t$.
A preliminary approximation $\Lambda^2_0$ of the appropriate amplitude $A^2$ is given by
\begin{equation}
    A^2 \approx \Lambda^2_0 \equiv -\frac{\sigma_{\rm{diff}}}{\sigma_{\rm{adv}}} \label{EQ:amp_approx}
\end{equation}
where the square in $A^2$ and $\Lambda^2_0$ is due to the quadratic nonlinearity of the eigenfunctions in the advective term and the minus sign is due to the fact that $\sigma_{\rm{adv}} < 0$ and $\sigma_{\rm{diff}} > 0$ in general.
We illustrate these trends in Figure~\ref{fig:iteration_spectra}. 
The subscript 0 denotes a preliminary estimate which we later refine.
In general, when $\Lambda^2_0 > 0$, the root-finding algorithm is stable.
The alternative can occur when the timestep approaches machine precision, in which case $\sigma_{\rm{diff}}$ and $\sigma_{\rm{adv}}$ are subject to floating point errors of comparable order to themselves.

Using the approximate amplitude $\Lambda_0^2$, the background temperature $\overline{T}(z, t_0)$ is then evolved to $t=t_0+\Delta t$ according to \eq{EQ:T0_IVP} and another eigenvalue solve is performed. 
This background temperature is typically close to, but not exactly, marginally stable.
We use Newton's method to find an amplitude $A^2$ such that the maximum growth rate is zero to within $10^{-9}$.
Marginally-stable modes do not oscillate in time, i.e.~$\sigma = 0$ implies $\omega = 0$.
This agrees with the conventional notion of exchange of stabilities \cite{drazin_reid_2004}.
Crucially, we do not assume the $k_x$ of the marginally-stable mode is fixed.
In Section~\ref{sec:multiple_modes} we specify procedures for the treatment of multiple simultaneously marginal modes.
